VIPRE: Virtualized Intermittently-Powered Runtime Environment

### 2017-07-10 paper/intro.tex
Second pass on the Intro ready

Two recent programming and execution model~\cite{alpaca,chain} proposed {\em task-based} paradigm, that require the programmer to statically decompose their program into tasks. A task resembles a function with no callers. Task can include arbitrary computation and, upon completion, is guaranteed to have executed {\em atomically} and to have logically {\em committed} memory updates. The programmer expresses the flow of control from one task to another and task control-flow may be input-dependent~\cite[Fig. 4]{chain}. An important programming challenge in task-based models is that the length of execution of a task is limited by the total amount of energy that a device can buffer. Assuming that input power is negligible compared to operating power, a task will never be able to complete if its execution consumes more energy than the system can buffer. Transitioning from one task to the next imposes a cost and excessively small tasks may be inefficient. Appropriately sizing tasks to ensure progress and high performance is a key challenge for task-based intermittent systems.

In this work, we develop \sys: a new programming and execution model that uses two new runtime features: (i) {\em volatile memory virtualization} to leverage efficient volatile memory and (ii) {\em dynamic task coalescing} to efficiently execute tasks without exceeding available energy. \sys's memory virtualization mechanism uses a device's SRAM as the working memory during a task's execution, populating the SRAM with a page of data from FRAM as necessary at the start of the task. At the end of the task, the SRAM is committed back to FRAM using efficient DMA block copies~\cite{kistler_micro_2006}. When a task accesses data outside of a page in SRAM, the task must fetch data from FRAM.

@@

With this paper we provide the following list of contributions.

\begin{enumerate}

\item {\bf Evaluation of data swapping strategies:} We propose and explore two data swapping strategies for intermittently-powered systems based on FRAM/SRAM architecture. The first strategy is {\em demand paging}, which swaps the SRAM page with a new page from FRAM, buffering the swapped-out page until commit. The second strategy is {\em buffered direct access}, which directly reads and writes FRAM relying on dynamic double-buffering to ensure memory is consistent at
commit.

\todo{Once paging is fully implemented--update the paragraph}{Przemek}
\todo{Provide numerical results for this section}{Przemek}
\todo{Find better name for demand paging}{Brandon}

\item {\bf Task coalescing mechanism:} \sys's core contribution is the introduction of dynamic task coalescing mechanism. It allows the programmer or a compiler to intuitively decompose the program into small tasks that amortize fixed per-task overheads, yet present no risk of exceeding device energy capacity. As such a decomposition executes, \sys {\em coalesces} dynamically consecutive tasks. Coalescing elides the commits of coalesced tasks by buffering multiple tasks' updates in an FRAM commit buffer. Periodically, as the span of the coalesced task grows, \sys ends coalescing and commits the state of the coalesced task. If a power failure interrupts a sequence of coalesced tasks, \sys adaptively reduces the number of tasks in that sequence that it will
coalesce, committing sooner in future executions. Consequently, \sys's coalescing mechanism allows a program to execute efficiently across a range of
energy buffer sizes, avoiding transition overheads in larger buffers, and
ensuring progress in smaller buffers.

\todo{Provide numerical results for this section}{Przemek}

\item {\bf Extensive \sys evaluation:} We evaluated \sys on a collection of benchmark programs~\cite[Sec. 5]{chain}, and compare directly to state-of-the art task-based runtime~\cite{chain}, showing that \sys has high performance and can flexible target a variety of platforms without recompiling the code. We further demonstrate the efficiency and versatility of \sys with an end-to-end evaluation of a sensing and data processing application deployed in a lab environment.

\todo{add detailed text about evaluation results}{Przemek}

\end{enumerate}

>>>>

Two recent works~\cite{alpaca,chain} proposed {\em task-based} programming and execution paradigm, that require a programmer to statically decompose a program into tasks. A task resembles a function with no callers. Task can include arbitrary computation and, upon completion, is guaranteed to have executed {\em atomically} and to have logically {\em committed} memory updates. The programmer expresses the flow of control from one task to another and task control-flow may be input-dependent~\cite[Fig. 4]{chain}. An important programming challenge in task-based models is that the length of execution of a task is limited by the total amount of energy that a device can buffer and {\em does not change after compilation}. Assuming that input power is negligible compared to operating power, a task will never be able to complete if its execution consumes more energy than the system can buffer. On the other hand, transitioning from one task to the next imposes a cost and excessively small tasks may be inefficient.

This brings the following research question: {\bf how to approach the appropriate task size to ensure progress and high performance given an unknown energy buffer size and energy arrival process?} More specifically, we ask two questions: (i) how to eliminate the need for hardware support while enabling the intermittent execution model to dynamically adapt task size {\em after compilation}, and (ii) how to reduce the energy consumption caused by frequent memory swapping operations during task traversals?

To answer these, in this work we develop {\bf \sys}: a new programming and execution model that utilizes two new runtime features: {\em dynamic task coalescing} to efficiently execute tasks without exceeding available energy (answering question (i)) and  {\em volatile memory virtualization} to leverage efficient volatile memory (answering question (ii)). \sys's novel memory virtualization mechanism uses a device's SRAM as the working memory during a task's execution, populating the SRAM with a page of data from FRAM as necessary at the start of the task. At the end of the task, the SRAM is committed back to FRAM using efficient DMA block copies~\cite{kistler_micro_2006}. When a task accesses data outside of a page in SRAM, the task must fetch data from FRAM.

@@

With this paper we provide the following list of contributions.
%
\begin{enumerate}
%
\item {\bf Task coalescing mechanism:} \sys's core contribution is the introduction of dynamic task coalescing mechanism. It allows the programmer or a compiler to intuitively decompose the program into small tasks that amortize fixed per-task overheads, yet present no risk of exceeding device energy capacity. As such a decomposition executes, \sys {\em coalesces} dynamically consecutive tasks. Coalescing elides the commits of coalesced tasks by buffering multiple tasks' updates in an FRAM commit buffer. Periodically, as the span of the coalesced task grows, \sys ends coalescing and commits the state of the coalesced task. If a power failure interrupts a sequence of coalesced tasks, \sys adaptively reduces the number of tasks in that sequence that it will
coalesce, committing sooner in future executions. Consequently, \sys's coalescing mechanism allows a program to execute efficiently across a range of
energy buffer sizes, avoiding transition overheads in larger buffers, and
ensuring progress in smaller buffers.
%
\todo{Provide numerical results for this section}{Przemek}
%
\item {\bf Evaluation of data swapping strategies:} We propose and explore two data swapping strategies for intermittently-powered systems based on FRAM/SRAM architecture. The first strategy is {\em demand paging}, which swaps the SRAM page with a new page from FRAM, buffering the swapped-out page until commit. The second strategy is {\em buffered direct access}, which directly reads and writes FRAM relying on dynamic double-buffering to ensure memory is consistent at
commit. 
%
\todo{Once paging is fully implemented--update the paragraph}{Przemek}
\todo{Provide numerical results for this section}{Przemek}
\todo{Find better name for demand paging}{Brandon}
%
\item {\bf Extensive \sys evaluation:} We evaluated \sys on a collection of benchmark programs~\cite[Sec. 5]{chain}, and compare directly to state-of-the art task-based runtime~\cite{chain}, showing that \sys has high performance and can flexible target a variety of platforms without recompiling the code. We further demonstrate the efficiency and versatility of \sys with an end-to-end evaluation of a sensing and data processing application deployed in a lab environment. 
%
\todo{add detailed text about evaluation results}{Przemek}
%
\end{enumerate}

### 2017-07-10 paper/paper.tex
Paragraph 1 revised with 10 new references added
Second pass on the Intro ready

\newcommand{\TODO}[1]{{\bf \em \textcolor{red}{TODO: #1}\xspace}}
\newcommand{\sys}{VIPRE\xspace}
\newcommand{\sysfull}{Virtualized Intermittently-Powered Runtime Environment\xspace}

@@

\subsection{Problem Statement}
How to approach the optimal task size given an unknown energy buffer size and environment conditions? 
	 \begin{enumerate}
		 \item How to eliminate the need for the hardware support while enabling the Modular Intermittent Execution model to dynamically adapt its task size?
		 \item How to enable code portability under the Modular Intermittent execution model? 
		 \item How to reduce the energy consumption of executing a task? 
	\end{enumerate}

>>>>

\newcommand{\TODO}[1]{{\bf \em \textcolor{red}{TODO: #1}\xspace}}

\newcommand{\todo}[2]{\textcolor{red}{$\bullet$ \textbf{Task:} #1; \textbf{Responsible:} #2}}

\newcommand{\sys}{VIPRE\xspace}
\newcommand{\sysfull}{Virtualized Intermittently-Powered Runtime Environment\xspace}

@@
